---
title: "Observation level random effects"
author: "Thierry Onkelinx"
date: "27 juni 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  autodep = TRUE
)
```

## Definition

An observation level random effect (OLRE) is a random effect with a different level for each observation. Combined with a Poisson distribution is can capture moderate overdispersion.

## Example

The model below is a simple mixed model. $\delta_{ijk}$ is the ORLE.

$$Y_{ijk} \sim Poisson(\mu_{ijk})$$

$$\log(\mu_{ijk}) = \eta_{ijk}$$

$$\eta_{ijk} = \beta_0 + \beta_1 X_i + b_j + \delta_{ijk}$$

$$b_j \sim N(0, \sigma_b)$$

$$\delta_{ijk} \sim N(0, \sigma_{olre})$$

Let's generate some overdispersed data using a negative binomial distribution.

```{r echo = TRUE}
set.seed(324)
n.i <- 10
n.j <- 10
n.k <- 10
beta.0 <- 1
beta.1 <- 0.3
sigma.b <- 0.5
theta <- 5
dataset <- expand.grid(
  X = seq_len(n.i),
  b = seq_len(n.j),
  Replicate = seq_len(n.k)
)
rf.b <- rnorm(n.j, mean = 0, sd = sigma.b)
dataset$eta <- beta.0 + beta.1 * dataset$X + rf.b[dataset$b]
dataset$mu <- exp(dataset$eta)
dataset$Y <- rnbinom(nrow(dataset), mu = dataset$mu, size = theta)
dataset$OLRE <- seq_len(nrow(dataset))
```

```{r}
library(ggplot2)
ggplot(dataset, aes(x = X, y = Y, group = Replicate)) +
  geom_line() +
  geom_point() +
  facet_wrap(~b)
```

Next we fit the model with an observation level random effect.

```{r}
library(lme4)
m1 <- glmer(Y ~ X + (1 | b) + (1 | OLRE), data = dataset, family = poisson)
summary(m1)
```

The estimates for $\beta_0$, $\beta_1$ and $\sigma_b$ are reasonably close to the ones used to generate the data.

## Check the sanity of the model

Models with an OLRE should be used carefully. Because OLRE can have a very strong influence on the model. One should always check the standard deviation of the ORLE. High standard deviations are a good indication for problems. Let's show why.

The main model (without OLRE) is $\gamma_{ijk} = \beta_0 + \beta_1 X_i + b_j$. The OLRE correct this for overdispersion $\eta_{ijk} = \gamma_{ijk} + \delta_{ijk}$. Since $\delta_{ijk} \sim N(0, \sigma_{olre})$, the 2.5% and 97.5% quantiles of this distribution define a range in which 95% of the plausible OLRE values are. The 2.5% quantile ($\delta_{lcl}$) indicates a strong but plausible downward correction, the 97.5% quantile  ($\delta_{ucl}$) an upward correction. So due to the corrections we have $\eta_{ijk} = (\gamma_{ijk} + \delta_{lcl}; \gamma_{ijk} + \delta_{ucl}) = \gamma_{ijk} + (\delta_{lcl};\delta_{ucl})$.In a Poisson distribution with log-link we have $\log(\mu_{ijk}) = \eta_{ijk}$ or $\mu_{ijk} = e ^{\eta_{ijk}} = e ^{\gamma_{ijk} + (\delta_{lcl};\delta_{ucl})}$. The ratio $r_{ijk}$ between the upper and lower boundary of $\mu_{ijk}$ is:

$$r_{ijk} = \frac{e ^{\gamma_{ijk} + \delta_{ucl}}}{e ^{\gamma_{ijk} + \delta_{lcl})}}$$

$$\log(r_{ijk}) = \log(\frac{e ^{\gamma_{ijk} + \delta_{ucl}}}{e ^{\gamma_{ijk} + \delta_{lcl})}})$$

$$\log(r_{ijk}) = \log(e ^{\gamma_{ijk} + \delta_{ucl}}) - \log(e ^{\gamma_{ijk} + \delta_{lcl}))}$$

$$\log(r_{ijk}) = \gamma_{ijk} + \delta_{ucl} - \gamma_{ijk} - \delta_{lcl} = \delta_{ucl} - \delta_{lcl}$$

The 2.5% quantile of $N(0, \sigma_{olre})$ is $\delta_{lcl} = -1.96 \sigma_{olre}$. The 97.5% quantile of $N(0, \sigma_{olre})$ is $\delta_{lcl} = +1.96 \sigma_{olre}$. Hence

$$\log(r_{ijk}) = 1.96 * \sigma_{olre} - (-1.96\sigma_{olre}) \approx 4 \sigma_{olre}$$

Or

$$r_{ijk} \approx 50.3 e ^ {\sigma_{olre}}$$

So when two observations have the same fixed and random covariates, expect of the OLRE, then the ratio between a fitted value with high and low OLRE is a whopping 50 times $e ^ {\sigma_{olre}}$.

This can be reasonable for small values of $\sigma_{olre}$.

```{r}
ratio <- data.frame(SigmaORLE = seq(0, 2.5, length = 101))
multiplier <- exp(diff(qnorm(c(.025, 0.975))))
ratio$Ratio <- multiplier * exp(ratio$SigmaORLE)
ggplot(ratio, aes(x = SigmaORLE, y = Ratio)) +
  geom_line()
```

But it goes though the roof quickly with even moderate values of $\sigma_{olre}$.

```{r}
ratio <- data.frame(SigmaORLE = seq(0, 10, length = 101))
multiplier <- exp(diff(qnorm(c(.025, 0.975))))
ratio$Ratio <- multiplier * exp(ratio$SigmaORLE)
ggplot(ratio, aes(x = SigmaORLE, y = Ratio)) +
  geom_line()
```

## Example with strong overdispersion

```{r echo = TRUE}
theta <- 0.15
dataset$Y2 <- rnbinom(nrow(dataset), mu = dataset$mu, size = theta)
```

```{r}
ggplot(dataset, aes(x = X, y = Y2, group = Replicate)) +
  geom_line() +
  geom_point() +
  facet_wrap(~b)
```

```{r echo = TRUE}
m2 <- glmer(Y2 ~ X + (1 | b) + (1 | OLRE), data = dataset, family = poisson)
summary(m2)
m2.nb <- glmer.nb(Y2 ~ X + (1 | b), data = dataset)
summary(m2.nb)
```

The estimates for $\beta_0$, $\beta_1$ and $\sigma_b$ are quite different with the ORLE model. The negative binomial model performs reasonable for $\beta_0$, $\beta_1$ but not for $\sigma_b$.
